Reasoning Graphically about Quantum Computation

### 2008-02-27 papers/reasoning-graphically/paper.tex
section on grph reps done
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1377 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

@@

\section{Compact Closed Categories and Graphs}
\label{sec:mono-categ-graphs}

@@

A \emph{directed graph}\footnote{
Equivalently:  a directed graph is a functor $G$ from  $\bullet
\pile{\rTo\\\rTo} \bullet$ to \catSet;  a graph morphism is then a natural
transformation  $f: G \Rightarrow H$.
} consists of a 4-tuple $(V,E,s,t)$ where $V$
and $E$ are sets, respectively of \emph{vertices}\footnote{We will use
  the words ``vertex'' and ``node'' interchangeably.} and \emph{edges},
and $s$ and $t$ are maps 
\begin{diagram}
  E & \pile{\rTo^{\qquad s\qquad}\\\rTo_t} & V
\end{diagram}
which we call \emph{source} and \emph{target}.  The \emph{in-degree}
of a vertex $v$ is the cardinality of $t^{-1}(v)$; its out-degree is
the cardinality of $s^{-1}(v)$.  The \emph{degree} of a vertex is
the sum of its in- and out-degrees.

@@

Augmented by some additional structure, graphs form a compact closed
category.  We present an overview of this structure:  the details
omitted here can be found in \cite{Duncan:thesis:2006}.

A \emph{concrete graph} $\Gamma$ is 5-tuple $(G, \dom\Gamma, \cod\Gamma,
<_{\text{in}(\cdot)}, <_{\text{out}(\cdot)})$ where:
\begin{itemize}
\item $G = (V,E,C,s,t)$ is a graph;
\item $\dom\Gamma$ and $\cod\Gamma$ are totally ordered disjoint of
  degree one vertices of $G$.
\item $<_{\text{in}(\cdot)}$ is a family of maps, indexed by $V$ such
  that $<_{\text{in}(v)} : \text{in}(v) \rTo^\isomorphism
  \mathbb{N}_k$ where $k = \sizeof{\text{in}(v)}$.
\item $<_{\text{out}(\cdot)}$ is a family of maps, indexed by $V$ such
  that $<_{\text{out}(v)} : \text{out}(v) \rTo^\isomorphism
  \mathbb{N}_{k'}$ where $k' = \sizeof{\text{out}(v)}$.
\end{itemize}

Since the sets $\dom\Gamma$ and $\cod\Gamma$ consist of vertices of degree
one, we can assign a polarity to each one:  $v \mapsto +$ if the edge
incident at $v$ is an incoming edge; $v \mapsto -$ otherwise.  Hence
$\cod \Gamma$ and $\dom \Gamma$ are \emph{signed sets}.  Concrete
graphs form a category whose objects are

can be seen as giving a signing to each element of the ordinal
one, we can view these sets as maps from the ordinal

can be viewed as a maps from an

representation of freely generated compact closed  categories  as
graphs;

need to define embedding and subgraph

theorem: two arrows in a a freely generated compact closed  category
are equal iff and only if their graphs are equal.

Note that the notion of vertex and graph have the same external
structure so it is possible to view a subgraph as a vertex and vice versa.

@@

To model the behaviour of quantum systems certain additional equations
must be satisfied.  Each family $(\delta, \epsilon)$ must form a
\emph{classical structure}
\cite{Coecke2006Quantum-Measure,Coecke2006POVMs-and-Naima}:  the pair
$(\delta, \epsilon)$  should form a cocommuative comonoid, the pair
$(\delta^\dag,\epsilon^\dag)$ must form a commutative monoid, and
together they must satisfy the isometry and frobenius equations. In
addition the family $\alpha$ must form an abelian group with $\alpha^\dag
= -\alpha$.  These equations are presented graphically below for the
green family;  the red vertices obey all the same equations
\begin{description}
\item[Comonoid Laws] 
\[
\begin{array}{ccccccccccccccc}
  \inlinegraphic{3em}{comonoid-assoc1}  &=\;&
  \inlinegraphic{3em}{comonoid-assoc2}
  &\qquad\qquad&
  \inlinegraphic{3em}{comonoid-unit1}  &=\;&
  \inlinegraphic{3em}{comonoid-unit2}  &=\;&
  \inlinegraphic{3em}{comonoid-unit3} 
  &\qquad\qquad&
  \inlinegraphic{3em}{comonoid-comm1}  &=\;&
  \inlinegraphic{3em}{comonoid-comm2}
\end{array}
\]
\item[Monoid Laws] 
\[
\begin{array}{ccccccccccccccc}
  \inlinegraphic{3em}{monoid-assoc1}  &=\;&
  \inlinegraphic{3em}{monoid-assoc2}
  &\qquad\qquad&
  \inlinegraphic{3em}{monoid-unit1}  &=\;&
  \inlinegraphic{3em}{monoid-unit2}  &=\;&
  \inlinegraphic{3em}{monoid-unit3} 
  &\qquad\qquad&
  \inlinegraphic{3em}{monoid-comm1}  &=\;&
  \inlinegraphic{3em}{monoid-comm2}
\end{array}
\]
\item[Isometry]
\[
\begin{array}{ccc}
  \inlinegraphic{3.5em}{isometry1}  &=\;&
  \inlinegraphic{3.5em}{isometry2}  
\end{array}
\]
\item[Frobenius]
  \[
  \begin{array}{ccccc}
    \inlinegraphic{3em}{frobenius1}  &=\;&
    \inlinegraphic{3em}{frobenius2} &=\;&
    \inlinegraphic{3em}{frobenius3}
  \end{array}
  \]
\item[Compact Structure] 
  \[
  \begin{array}{ccccccc}
    \inlinegraphic{2.5em}{compact1}  &=\;&
    \inlinegraphic{1.3em}{compact2}  
    &\qquad\qquad&
    \inlinegraphic{2.5em}{compact3}  &=\;&
    \inlinegraphic{1.3em}{compact4}  
  \end{array}
  \]
  \item[Abelian Unitary Group]
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{2.2em}{group1}  &:=\;&
      \inlinegraphic{2.2em}{group2}  &=\;&
      \inlinegraphic{2.2em}{group3} 
      &\qquad\qquad&
      \left( \inlinegraphic{2.2em}{group4}\!\right)^\dag  &=\;&
      \inlinegraphic{2.2em}{group5} 
      &\qquad\qquad&
      \inlinegraphic{3.5em}{group6}  &=\;&
      \inlinegraphic{3.5em}{group7}  &=\;&
      \inlinegraphic{3.5em}{group8} 
    \end{array}
    \]
  \item[Bilinearity] 
    \[
    \begin{array}{cccccccccccc}
      \inlinegraphic{3.5em}{alpha-commute1}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute2}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute3} 
      &\qquad\qquad&
      \inlinegraphic{3.5em}{alpha-commute4}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute5}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute6} 
    \end{array}
    \]
\end{description}
Additionally, the two families satisfy equations which make them a
\emph{scaled bialgebra}.
\begin{description}
\item[Bialgebra Laws] Let  $\inlinegraphic{1em}{bialgebra1} := 
      \inlinegraphic{1.5em}{bialgebra2}$;  then:
  \[
  \begin{array}{ccccccccccccccc}
      \inlinegraphic{3.5em}{bialgebra3}  &=\;&
      \inlinegraphic{3em}{bialgebra4}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra5}  &=\;&
      \inlinegraphic{1.5em}{bialgebra6}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra7}  &=\;&
      \inlinegraphic{1.5em}{bialgebra8} \\ \\
      \inlinegraphic{3.5em}{bialgebra9}  &=\;&
      \inlinegraphic{3em}{bialgebra10}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra11}  &=\;&
      \inlinegraphic{1.5em}{bialgebra12}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra13}  &=\;&
      \inlinegraphic{1.5em}{bialgebra14} 
  \end{array}
  \]
\end{description}

>>>>

History of diagrammatic representation going back to ???

@@

\section{Graphs and Compact Closed  Categories  }
\label{sec:mono-categ-graphs}

@@

A \emph{directed graph}\footnote{
Equivalently:  a directed graph is a functor $G$ from  $\bullet
\pile{\rTo\\\rTo} \bullet$ to \catSet;  a graph morphism is then a natural
transformation  $f: G \Rightarrow H$.
} consists of a 4-tuple $(V,E,s,t)$ where $V$
and $E$ are sets, respectively of \emph{vertices}\footnote{We will use
  the words ``vertex'' and ``node'' interchangeably.} and \emph{edges},
and $s$ and $t$ are maps 
\begin{diagram}
  E & \pile{\rTo^{\qquad s\qquad}\\\rTo_t} & V
\end{diagram}
which we call \emph{source} and \emph{target}.  Define sets
$\text{in}(v) = t^{-1}(v)$ and $\text{out}(v) = s^{-1}(v)$ of \emph{incoming}
and \emph{outgoing} edges at a vertex $v$.  
The \emph{degree} of a vertex $v$ is $\sizeof{\text{in}(v)} +
\sizeof{\text{out}(v)}$.

@@

Augmented by some additional structure, graphs form a compact closed
category.  In section \ref{sec:graph-repr-comp} we will describe this
structure, but first we review the concept of compact closed  category.

\subsection{Compact Closed Categories}
\label{sec:comp-clos-categ}

\subsection{Graph Representations for Compact Closed Categories}
\label{sec:graph-repr-comp}

Graphs with certain additional structure give a
representation for compact closed  categories; we now give an overview
of this construction.  The details omitted here can be found in
\cite{Duncan:thesis:2006}.

A \emph{concrete graph} $\Gamma$ is 5-tuple $(G, \dom\Gamma, \cod\Gamma,
<_{\text{in}(\cdot)}, <_{\text{out}(\cdot)})$ where:
\begin{itemize}
\item $G = (V,E,s,t)$ is a graph;
\item $\dom\Gamma$ and $\cod\Gamma$ are totally ordered disjoint sets of
  degree one vertices of $G$.
\item $<_{\text{in}(\cdot)}$ is a family of maps, indexed by $V$ such
  that $<_{\text{in}(v)} : \text{in}(v) \rTo^\isomorphism
  \mathbb{N}_k$ where $k = \sizeof{\text{in}(v)}$.
\item $<_{\text{out}(\cdot)}$ is a family of maps, indexed by $V$ such
  that $<_{\text{out}(v)} : \text{out}(v) \rTo^\isomorphism
  \mathbb{N}_{k'}$ where $k' = \sizeof{\text{out}(v)}$.
\end{itemize}

Since the sets $\dom\Gamma$ and $\cod\Gamma$ consist of vertices of degree
one, we can assign a polarity to each one:  $v \mapsto +$ if the edge
incident at $v$ is an incoming edge; $v \mapsto -$ otherwise.  Hence
$\cod \Gamma$ and $\dom \Gamma$ are \emph{ordered signed sets}.  Given any
ordered signed set $S$ we write $S^*$ for the same ordered set, but
the opposite signing.   Given two such sets we can define their disjoint
union $R+S$ as the disjoint union of the underlying sets, inheriting
the signing and the order from $R$ and $S$, with the convention that
$r < s$ for all $r\in R, s\in S$.

\begin{proposition}
Concrete graphs form a compact closed category whose objects are
ordered signed sets and whose arrows $f:A\to B$ are concrete graphs with
$\cod f = B$  and $\dom f = A^*$.
\end{proposition}
For each ordered signed set $A$, the identity map for $A$ has
$\dom \id{A} = A^*$ and $\cod \id{A} = A$; its underlying
graph has $E = A$ and $V = A^* + A$ with $t(a) = a$ and $s(a) =
a^*$.  Given a pair of concrete graphs $f:A\to B$ and $g:B\to C$ their
composition $g\circ f:A\to C$ is constructed by merging the two
graphs, erasing the vertices of $\cod f$ and $\dom g$, and identifying
the edges previously incident at the deleted vertices.  (Due to the
opposite polarity of the domain and codomain the edges have compatible
direction.)  The tensor product on objects $A,B$ is simply $A+B$;
given $f:A\to B$, $g: C\to D$, the graph of $f \otimes g$ is the
disjoint union of the graphs of $f$ and $g$.  The unit for the tensor
is the empty set.  The morphisms $d : I \to A \otimes A^*$,
$e: A^* \otimes A \to I$ have the same underlying graph, but $\dom d =
\emptyset$, $\cod d = A+A^*$, $\dom e = A^*+A$ and $\cod e = \emptyset$.

This category captures exactly the axioms for compact closed
structure, in the sense that any freely generated compact closed
category can be represented by concrete graphs.  We will consider
a collection of basic terms\footnote{See \cite{Duncan:thesis:2006} for
  a more thorough description of the nature of the terms here} $F$
whose types are vectors of some set of basic types $T$.  Then:

\begin{definition}
  A \emph{$T,F$-labelling} $\theta$ for a concrete graph $\Gamma$ is a pair of
  maps  $\theta_T : E \to T$ and $\theta_F : (V - \cod\Gamma -
  \dom\Gamma) \to F$  such that for each vertex  $v$, if
  $\text{in}(v) = \langle a_1, \ldots, a_n\rangle$ and $\text{out}(v)
  = \langle b_1, \ldots, b_m\rangle$ then 
  \[
  \theta v : \langle \theta a_1, \ldots, \theta a_n \rangle
  \to 
  \langle \theta b_1, \ldots, \theta b_m \rangle
  \]
  Call a concrete graph $\Gamma$ \emph{$T,F$-labellable} if there exists an 
  $T,F$-labelling for it; if $\theta$ is a labelling for $\Gamma$, then
 the pair $(\Gamma,\theta)$ is an \emph{$T,F$-labelled graph}.
\end{definition}

The $T,F$-labelled graphs form a compact closed category in the same
way as the the concrete graphs, subject to the further restriction
that arrows are composable only when their labellings agree.

\begin{theorem}[\cite{Duncan:thesis:2006}]
  Let \catC be a compact closed category  freely generated by some set
  of arrows $F$ and ground types $T$;  then \catC is equivalent to the
  category  of $T,F$-labelled graphs.
\end{theorem}

Given a compact closed  category  \catC generated by some basic set of
operations,  the arrows of \catC have a canonical representation as
labelled graphs.  A consequence of the theorem is then that two arrows
are equal by the equations of the compact closed  structure if and
only if their graph representations are equal.

As a final remark before moving on, note that the external structre of vertex
in a concrete graph is esentially the same as that of a complete
graph;  hence one can consistently view subgraphs as vertices, and
abstract over the their internal structure.

@@

To model the behaviour of quantum systems certain additional equations
must be satisfied.  Each family $(\delta, \epsilon)$ must form a
\emph{classical structure}
\cite{Coecke2006Quantum-Measure,Coecke2006POVMs-and-Naima}:  the pair
$(\delta, \epsilon)$  should form a cocommuative comonoid, the pair
$(\delta^\dag,\epsilon^\dag)$ must form a commutative monoid, and
together they must satisfy the isometry and frobenius equations. In
addition the family $\alpha$ must form an abelian group with $\alpha^\dag
= -\alpha$.  These equations are presented graphically below for the
green family;  the red vertices obey all the same equations
\begin{description}
\item[Comonoid Laws] 
\[
\begin{array}{ccccccccccccccc}
  \inlinegraphic{3em}{comonoid-assoc1}  &=\;&
  \inlinegraphic{3em}{comonoid-assoc2}
  &\qquad\qquad&
  \inlinegraphic{3em}{comonoid-unit1}  &=\;&
  \inlinegraphic{3em}{comonoid-unit2}  &=\;&
  \inlinegraphic{3em}{comonoid-unit3} 
  &\qquad\qquad&
  \inlinegraphic{3em}{comonoid-comm1}  &=\;&
  \inlinegraphic{3em}{comonoid-comm2}
\end{array}
\]
\item[Monoid Laws] 
\[
\begin{array}{ccccccccccccccc}
  \inlinegraphic{3em}{monoid-assoc1}  &=\;&
  \inlinegraphic{3em}{monoid-assoc2}
  &\qquad\qquad&
  \inlinegraphic{3em}{monoid-unit1}  &=\;&
  \inlinegraphic{3em}{monoid-unit2}  &=\;&
  \inlinegraphic{3em}{monoid-unit3} 
  &\qquad\qquad&
  \inlinegraphic{3em}{monoid-comm1}  &=\;&
  \inlinegraphic{3em}{monoid-comm2}
\end{array}
\]
\item[Isometry]
\[
\begin{array}{ccc}
  \inlinegraphic{3.5em}{isometry1}  &=\;&
  \inlinegraphic{3.5em}{isometry2}  
\end{array}
\]
\item[Frobenius]
  \[
  \begin{array}{ccccc}
    \inlinegraphic{3em}{frobenius1}  &=\;&
    \inlinegraphic{3em}{frobenius2} &=\;&
    \inlinegraphic{3em}{frobenius3}
  \end{array}
  \]
\item[Compact Structure] 
  \[
  \begin{array}{ccccccc}
    \inlinegraphic{2.5em}{compact1}  &=\;&
    \inlinegraphic{1.3em}{compact2}  
    &\qquad\qquad&
    \inlinegraphic{2.5em}{compact3}  &=\;&
    \inlinegraphic{1.3em}{compact4}  
  \end{array}
  \]
  \item[Abelian Unitary Group]
    \[
    \begin{array}{ccccccccccccccc}
      \inlinegraphic{2.2em}{group1}  &:=\;&
      \inlinegraphic{2.2em}{group2}  &=\;&
      \inlinegraphic{2.2em}{group3} 
      &\qquad\qquad&
      \left( \inlinegraphic{2.2em}{group4}\!\right)^\dag  &=\;&
      \inlinegraphic{2.2em}{group5} 
      &\qquad\qquad&
      \inlinegraphic{3.5em}{group6}  &=\;&
      \inlinegraphic{3.5em}{group7}  &=\;&
      \inlinegraphic{3.5em}{group8} 
    \end{array}
    \]
  \item[Bilinearity] 
    \[
    \begin{array}{cccccccccccc}
      \inlinegraphic{3.5em}{alpha-commute1}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute2}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute3} 
      &\qquad\qquad&
      \inlinegraphic{3.5em}{alpha-commute4}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute5}  &=\;&
      \inlinegraphic{3.5em}{alpha-commute6} 
    \end{array}
    \]
\end{description}
Additionally, the two families satisfy equations which make them a
\emph{scaled bialgebra}.
\begin{description}
\item[Bialgebra Laws] Let  $\inlinegraphic{1em}{bialgebra1} := 
      \inlinegraphic{1.5em}{bialgebra2}$;  then:
  \[
  \begin{array}{ccccccccccccccc}
      \inlinegraphic{3.5em}{bialgebra3}  &=\;&
      \inlinegraphic{3em}{bialgebra4}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra5}  &=\;&
      \inlinegraphic{1.5em}{bialgebra6}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra7}  &=\;&
      \inlinegraphic{1.5em}{bialgebra8} \\ \\
      \inlinegraphic{3.5em}{bialgebra9}  &=\;&
      \inlinegraphic{3em}{bialgebra10}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra11}  &=\;&
      \inlinegraphic{1.5em}{bialgebra12}
      \qquad\qquad
      \inlinegraphic{2.5em}{bialgebra13}  &=\;&
      \inlinegraphic{1.5em}{bialgebra14} 
  \end{array}
  \]
\end{description}
Last group of equations

Statement and Discussion of spider law.
